# Red Wine Quality Exploration  
## Bart Leatham  
### 04/18/2017

Wine experts judge wines on a variety of categories in an attempt to quantify 
a wines quality.  One of the categories used is the 'balance' of a wine.  The 
balance consists of 4 measureable traits, sweetness, acidity, tannin and 
alcohol content. While personal taste will always prevail over a wine experts 
opinion, this investigation into a [dataset](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-
hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1492550811255000&usg=
AFQjCNGNw9JZHE_6c9hEJDIGpoLyC5MSkA) 
containing the judgments of at least three wine experts will attempt to 
quantify the relationships between the traits of a balanced wine and the 
perceived quality rating by the judges. This dataset consists of data related 
to red Portuguese "Vinho Verde" wine.  

#### Source of the dataset  
Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, 
Telmo Matos and Jose Reis (CVRVV) @ 2009  

#### Attribute Information  
   1 - fixed acidity (tartaric acid - g / dm^3)  
   2 - volatile acidity (acetic acid - g / dm^3)  
   3 - citric acid (g / dm^3)  
   4 - residual sugar (g / dm^3)  
   5 - chlorides (sodium chloride - g / dm^3  
   6 - free sulfur dioxide (mg / dm^3)  
   7 - total sulfur dioxide (mg / dm^3)  
   8 - density (g / cm^3)  
   9 - pH  
   10 - sulphates (potassium sulphate - g / dm3)  
   11 - alcohol (% by volume)  
   Output variable (based on sensory data):   
   12 - quality (score between 0 and 10)  
   
#### Attribute Descriptions:  
   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
   (do not evaporate readily)  

   2 - volatile acidity: the amount of acetic acid in wine, which at too high 
   of levels can lead to an unpleasant, vinegar taste  

   3 - citric acid: found in small quantities, citric acid can add 'freshness' 
   and flavor to wines  

   4 - residual sugar: the amount of sugar remaining after fermentation stops, 
   it's rare to find wines with less than 1 gram/liter and wines with greater 
   than 45 grams/liter are considered sweet  

   5 - chlorides: the amount of salt in the wine  

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
   growth and the oxidation of wine  

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
   concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of 
   wine  

   8 - density: the density of water is close to that of water depending on 
   the percent alcohol and sugar content  

   9 - pH: describes how acidic or basic a wine is on a scale from 
   0 (very acidic) to 14 (very basic); most wines are between 3-4 on the 
   pH scale  

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
   (S02) levels, wich acts as an antimicrobial and antioxidant  

   11 - alcohol: the percent alcohol content of the wine  

   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)  

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

#install.packages("ggplot2", dependencies = T) 
#install.packages("knitr", dependencies = T)
#install.packages("dplyr", dependencies = T)
#install.packages("Hmisc", dependencies = T)
#install.packages("GGally", dependencies = T)
#install.packages("memisc", dependencies = T)
#install.packages("corrplot")


library(ggplot2)
library(dplyr)
library(knitr)
library(GGally)
library(memisc)
library(corrplot)
library(gridExtra)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_data}
# Load the Data
wine_df <- read.csv('wineQualityReds.csv')

#remove the 'X' column
wine_df$X <- NULL

#create correlation matrix
wine_corr <- cor(wine_df)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Define_Functions}

#This is a sandbox for functions to be called throughout EDA

#function to create histogram of feature count
feature_hist <- function(feature, data, 
                         binwidth = diff(range(data[column]))/30) {
    return(ggplot(aes_string(x = feature), data = data) + 
             geom_bar())
  }

#function to create density plot of feature vs. quality_ranking
density_plot <- function(feature, data) {
  return(print(
    ggplot(aes_string(x = feature), data = data) +
      geom_density(aes(fill = Quality_Ranking), alpha = .3)
  ))
}

#function to create scatter plot of feature vs. quality_ranking
scatter_plot <- function(feature, data) {
  return(print(
    ggplot(aes_string(x = feature, y = "quality.bucket"), data = data) +
      geom_point(aes(fill = Quality_Ranking, color = Quality_Ranking ))
  ))
}

#function to create bar chart of feature vs quality
quality_bar <- function(feature, data) {
    return(print(ggplot(aes_string(x = "quality.bucket", y = feature), 
                        data = data) +
  geom_bar(stat = 'summary', fun.y = mean)
    ))
}

#function to create box plot of feature vs quality
quality_box <- function(feature, data, ylim = c(NA,NA)) {
  return(print(ggplot(aes_string(x = "Quality_Ranking", y = feature), 
                        data = data) +
                 scale_y_continuous(limit = ylim) +
                 geom_jitter( alpha = .3)  +
                 geom_boxplot(outlier.colour = "red", outlier.shape = 1,
                              alpha = .5, color = 'blue') +
                 stat_summary(fun.y = "mean", 
                 geom = "point", 
                 color = "red", 
                 shape = 8, 
                 size = 4)
  ))
}

#function to visualize outliers
outlier_plot <- function(feature, data) {
  return(ggplot(aes_string( x = 1, y = feature ), 
                      data = data) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red')) 
}
            
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Cleaning}
#create new quality.bucket with discrete "low", "mid", "high" buckets
wine_df$quality.bucket <- cut(wine_df$quality, c(3, 4, 6, 8), 
                              labels = c("low", "mid", "high"), 
                              include.lowest = TRUE)

#convert quality to a factor type
wine_df$quality <- factor(wine_df$quality)

#assign above to a variable for cleaner plotting
Quality_Ranking <- wine_df$quality.bucket

```
# Overview of the data
```{r echo=FALSE, message=FALSE, warning=FALSE, Structure}
str(wine_df)
summary(wine_df)
```
I converted the 'quality' feature to a factor as it is discrete values, and 
created a 'quality.bucket' variable with low, mid and high quality buckets for 
more meaningful analysis.  
It should be noted, as shown in the summary above, that there are many more 
datapoints falling in the 'mid'(1319) quality bucket than 'low'(63) or 
'high'(217).  This should be considered when making any assumptions about the 
analysis.

# Univariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('fixed.acidity', wine_df)
p2<- outlier_plot('fixed.acidity', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$fixed.acidity)
```
We see that fixed acidity has a close to normal distribution with a median of
7.90 and a mean of 8.32. The outlier boxplot reveals that values above ~12 are
statistical outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('volatile.acidity', wine_df)
p2<- outlier_plot('volatile.acidity', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$volatile.acidity)
```
Volatile acidity has a near normal distribution with a tail of 
outliers at higher values. The median and mean are close to the same, 
at .52 and .53 respectively. Values above 1 are shown to be statistical 
outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('citric.acid', wine_df)
p2<- outlier_plot('citric.acid', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$citric.acid)
```
Citric acid shows a very unusual spike at 0. According to tests conducted by 
Food Standards Australia sited [here](http://www.livestrong.com/article/189520-what-drinks-do-not-contain-
citric-acid/), many types of red wines do not contain any citric acid, so this
spike at 0 is not indicitave of an issue with the data.  The citric acid data
is shown to be quite clean, with minimal outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('residual.sugar', wine_df)
p2<- outlier_plot('residual.sugar', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$residual.sugar)
```

Residual sugar shows a normal distribution with a long tail of 
outliers.  The majority of the samples do fall within the lower range of all 
reported residual sugar values, with a median of 2.20 and mean of 2.54.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('chlorides', wine_df)
p2<- outlier_plot('chlorides', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$chlorides)
```

Chlorides content follows a similar distribution to that of residual sugar, 
though there is no physical relationship between them.  The tail of outliers is
significant.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('free.sulfur.dioxide', wine_df)
p2<- outlier_plot('free.sulfur.dioxide', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$free.sulfur.dioxide)
```
Free sulfur dioxide shows a normal distribution skewed right with a median of 
14.00 and a mean of 15.87, outliers are shown above a value of ~40.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('total.sulfur.dioxide', wine_df)
p2<- outlier_plot('total.sulfur.dioxide', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$total.sulfur.dioxide)
```
Total sulfur dioxide follows a similar distribution to free sulfur dioxide, 
this is not surprising as free sulfur dioxide is a subset of total sulfur 
dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('density', wine_df)
p2<- outlier_plot('density', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$density)
```
The density of the wine samples fall in a very narrow range of .99 to 1.0 with 
the median and mean essentially equal at 0.9968 and 0.9967 respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('pH', wine_df)
p2<- outlier_plot('pH', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$pH)
```
The pH of the wine samples has a normal distribution, with the median and mean
equal at 2 significant digits with a pH of 3.31.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('sulphates', wine_df)
p2<- outlier_plot('sulphates', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$sulphates)
```

Sulphates have a near normal distribution with a significant range of
outliers causing a tail at higher values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<- feature_hist('alcohol', wine_df)
p2<- outlier_plot('alcohol', wine_df)
grid.arrange(p1, p2, ncol = 2)
summary(wine_df$alcohol)
```

The median and mean alcohol content of the wine samples skews towards the lower
end of the range of the samples, with a median of 10.20 and mean of 10.42.

```{r echo=FALSE, message=FALSE, warning=FALSE}
feature_hist('quality', wine_df)
summary(wine_df$quality)

```
The histogram of wine quality highlights the fact that the majority of the 
wines in this dataset fall within the middle of the range of quality ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE}
feature_hist('quality.bucket', wine_df)
summary(wine_df$quality.bucket)
```

The quality of wines will be analyzed with the quality ranking assigned to the
buckets shown above.  Ratings of 3 & 4 are low, 5 & 6 are mid, 7 & 8 are high.


# Univariate Analysis

### What is the structure of your dataset?
1599 obs. of  13 variables.  11 input variables, 2 output variables.

### What is/are the main feature(s) of interest in your dataset?
This analysis is focused on the balance of the wines, of which the traits are 
sweetness, acidity, tannin and alcohol. The dataset does not contain a measure
for tannins, so the investigation will focus on sweetness 
(measured by residual.sugar), acidity (measured by pH), the alcohol content
and quality rating.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I am curious how the density of the wine will relate to the quality rating.

### Did you create any new variables from existing variables in the dataset?
I bucketed the quality ratings into a quality.bucket variable, with low, med, 
high being the buckets.  This allows for a more meaningful analysis, as the 
counts for some quality rankings were low compared to others.  

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?  
This dataset met tidy data requirements as delivered, so the only cleaning I 
performed was to remove the X column as it was just a line count index.  
Most of the features have an overall normal distribution, some skewed to the 
right, with a long tail of outliers.

# Bivariate Plots Section

### Correlation data
Below we see the correlation values between all input and output variables, as 
well as a graphical representation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlation}
#summary(wine_df)
head(round(wine_corr, 2), 13)
corrplot(wine_corr, method="circle", type="lower")
```

A quick look at the correlation between the variables shows that there are no 
strong correlations between the quality rating and the various input variables,
the strongest postive correlation being the alcohol content at r= 0.48, and the
strongest negative correlation being the volatile acidity at r= -0.39.


### Bivariate Plots
After perfoming multiple plotting techniques (hist, bar, density, scatter, box) 
for introductory analysis, I found that using a scatter and box plot of each 
variable vs. quality ranking yielded the most useful visual data for Bivariate 
analysis.  I am focusing my analysis on the previously mentioned 'balance' 
features of residual sugar, pH, alcohol content as well as the density.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
#iterate over features in wine_df and create initial exploratory plots 
quality_box('fixed.acidity', wine_df, c(4,14))
wc <- round(wine_corr[1, 12], 3)
print(paste0("The correlation between fixed.acidity and quality is ", wc))

```



```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('volatile.acidity', wine_df, c(0,1.2))
wc <- round(wine_corr[2, 12], 3)
print(paste0("The correlation between volatile.acidity and quality is ", wc))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('citric.acid', wine_df)
wc <- round(wine_corr[3, 12], 3)
print(paste0("The correlation between citric.acid and quality is ", wc))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

quality_box('residual.sugar', wine_df, c(1,4))

wc <- round(wine_corr[4, 12], 3)
print(paste0("The correlation between residual.sugar and quality is ", wc))

```

With a low correlation of r=0.014, residual sugar is not a driving factor in 
the quality rating of the wine samples.

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('chlorides', wine_df, c(0.025,.15))
wc <- round(wine_corr[5, 12], 3)
print(paste0("The correlation between chlorides and quality is ", wc))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('free.sulfur.dioxide', wine_df, c(0,50))
wc <- round(wine_corr[6, 12], 3)
print(paste0("The correlation between free.sulfur.dioxide and quality is ", 
             wc))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('total.sulfur.dioxide', wine_df, c(0,150))
wc <- round(wine_corr[7, 12], 3)
print(paste0("The correlation between total.sulfur.dioxide and quality is ",
             wc))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('density', wine_df)
wc <- round(wine_corr[8, 12], 3)
print(paste0("The correlation between density and quality is ", wc))
```
It appears that a lower density wine indicates that it will have a higher 
rating.  With the small range of measured density values, and the weak 
correlation of r=-0.175, the plot can be misleading, the affect of density 
on wine quality is not as strong as indicated by a quick look at the box plot.



```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('pH', wine_df)
wc <- round(wine_corr[9, 12], 3)
print(paste0("The correlation between pH and quality is ", wc))

```
Although the boxplot shows a trend of a lower pH yielding a higher quality
rating, with a correlation of r=-0.058, we cannot say that the pH significantly
impacts the quality rating of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('sulphates', wine_df, c(0.25,1.5))
wc <- round(wine_corr[10, 12], 3)
print(paste0("The correlation between sulphates and quality is ", wc))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
quality_box('alcohol', wine_df)
wc <- round(wine_corr[11, 12], 3)
print(paste0("The correlation between alcohol and quality is ", wc))

```

With a moderate correlation value of r=0.476, and little overlap of values
between the high quality and the mid or lower quality rankings, it can be 
stated that the alcohol content does affect the quality rating, higher alcohol
yields a higher rating.

## Bivariate Analysis
Upon analyzing the above plots, we can see that the higher the chloride and 
volatile.acidity, the lower the quality.  
The higher the fixed.acidity, sulfates, alcohol and citric.acid, the higher 
the quality.  
It is interesting to note, that if relying on the box plots alone, 
one could infer that the variables mentioned could have a large correlation to 
the quality of the wine. Since we have looked at the correlation values 
explicity, we know that this is not case, there is a correlation, though not a
strong one.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
Of the 4 features of interest, density and alcohol content have the strongest
correlations to the quality of the wine. Density has a r = -0.17 while alcohol
has r = 0.48 correlation to the quality of the wine. The pH and residual sugar
have low correlations (r = -0.06 and r = 0.01 respectively).  I expected
that the residual sugar would have more of an impact, as I often hear wines 
described as being too sweet.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
It was interesting to see that the correlation data reinforced some 
relationships that would be expected.  For instance, the various *.acid 
features have relatively strong correlations to the pH.

### What was the strongest relationship you found?
The strongest relationship that I found was between free.sulfur.dioxide and
total.sulfur.dioxide.  This makes sense, as the free.sulfur.dioxide is a subset
of the total.sulfur.dioxide, so it is not particularly interesting.

# Multivariate Plots Section
I was interested in looking deeper into the relationships between the residual 
sugar, density, pH, alcohol content and quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}

wc1 <- round(wine_corr[4, 11], 3)
wc2 <- round(wine_corr[9, 11], 3)
wc3 <- round(wine_corr[8, 11], 3)

ggplot(aes_string(x = 'residual.sugar', y = 'alcohol',
       color = Quality_Ranking), data = wine_df) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = 'seq', guide = guide_legend(title = 'Quality'), 
                     palette = 'Reds')

print(paste0("The correlation between residual.sugar and alcohol is ", wc1))
```
We can see that higher alcohol levels consistently yield higher quality 
ranking results.  The correlation between residual sugar and alcohol is very 
weak.  

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes_string(x = 'pH', y = 'alcohol',
       color = Quality_Ranking), data = wine_df) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = 'seq', guide = guide_legend(title = 'Quality'), 
                     palette = 'Reds')

print(paste0("The correlation between pH and alcohol is ", wc2))
```

The correlation between pH and alcohol is weak at r = 0.21.  This plot does
indicate again that higher alcohol content yields a more highly rated wine, and
that the pH range of highly rated wines is tighter than for mid or lower rated
wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes_string(x = 'density', y = 'alcohol',
       color = Quality_Ranking), data = wine_df) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(type = 'seq', guide = guide_legend(title = 'Quality'), 
                     palette = 'Reds')

print(paste0("The correlation between density and alcohol is ", wc3))
```

With a correlation between density and alcohol content of r=-0.496, we can see
that higher alcohol content yields a less dense wine.  This makes sense due to
the fact that alcohol is less dense than water, so the higher the alcohol 
content, the less dense the wine.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

For residual sugar, we can see that there are no high quality ranked wines with
a residual sugar level over ~9g/dm^3, giving credence to some wines being too
sweet.  
For the pH, we can see that above ~3.75 and below ~2.5 there are no high 
quality ranked wines, leading me to believe that a pH of 2.5-3.75 is the
range for a well balanced wine, not too acidic or basic.  It can be noted
though, that on the pH scale, that range is quite acidic.  
For the density of the wine, we cannot infer much as to quality, we do see
correlation to the fact that lower density results in higher alcohol levels, 
which we have seen have a moderate correlation to quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

ggplot(aes_string(x = "Quality_Ranking", y = "alcohol"), 
                        data = wine_df) +
                 geom_jitter( alpha = .3)  +
                 geom_boxplot(outlier.colour = "red", outlier.shape = 1,
                              alpha = .5, color = 'blue') +
  scale_y_continuous(limits = c(8,14)) +
   ggtitle("Wine Quality as a function of Alcohol") +
        labs(x = "Quality Ranking", y = "Alcohol %") +
                 stat_summary(fun.y = "mean", 
                 geom = "point", 
                 color = "red", 
                 shape = 8, 
                 size = 4)
```

### Description One
The alcohol content of the wine has been found to have the single largest 
affect on how highly the wine quality is ranked, with a correlation of r = 0.48.
It is clear from the plot above that wines with a higher alcohol content are 
rated more highly, while no such distinction can be made between the low and 
mid quality groups.  I am curious if with a larger dataset, we would see a drop 
off in quality ratings after the alcohol content hits a certain level.  

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes_string(x = "Quality_Ranking", y = "residual.sugar"), 
                        data = wine_df) +
                 geom_jitter( alpha = .3)  +
                 geom_boxplot(outlier.colour = "red", outlier.shape = 1,
                              alpha = .5, color = 'blue') +
   scale_y_continuous(limits = c(1,4)) +
   ggtitle("Wine Quality as a function of Residual Sugar") +
        labs(x = "Quality Ranking", y = "Residual Sugar (g/dm^3)") +
                 stat_summary(fun.y = "mean", 
                 geom = "point", 
                 color = "red", 
                 shape = 8, 
                 size = 4)
```

### Description Two
Residual sugar was not found to have an affect on how highly a wine is rated. A
correlation of r = 0.014 statistically validates this observation.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes_string(x = "Quality_Ranking", y = "pH"), 
                        data = wine_df) +
                 geom_jitter( alpha = .3)  +
                 geom_boxplot(outlier.colour = "red", outlier.shape = 1,
                              alpha = .5, color = 'blue') +
   ggtitle("Wine Quality as a function of pH") +
        labs(x = "Quality Ranking", y = "pH (measure of acidity)") +
                 stat_summary(fun.y = "mean", 
                 geom = "point", 
                 color = "red", 
                 shape = 8, 
                 size = 4)
```

### Description Three
There appears to be a trend of lower pH yeilding a higher wine quality rating,
though with a correlation of r=-0.06, we know that this is not the case with
any statistical significance.

# Reflection
This red wine dataset provided 1599 observations of 11 input and 1 output 
variable.  I first looked into the distribution of the variable values, then 
moved on to their correlations.  Once I saw that the correlations were 
generally weak, especially with regard to the reported quality rating of the 
wine, I realized that I needed to come up with a direction to take the 
investigation.  At that point I researched how wine quality is generally 
defined by experts, and decided to pursue the measure of balance for a wine
and investigated its traits as much as possible with the data provided.  
The main struggle I faced in this investigation was the fact that none of the
independant variables were statistically strongly correlated to the quality
ranking of the wines.  
I was surprised to see that the features were not more strongly correlated to
the quality rating of the wines. Before this investigation I would have 
guessed that alcohol, pH and sweetness all would have been more highly 
correlated with the quality rating. To me this proves that wine quality ratings
are highly subjective, and a wine consumer would do best to trust their
tastebuds (and wallets) over being overly concerned with how a wine expert
views their choice.
For future work with this dataset I would be interested in having both a 
measure of tannins, and the price of the wine included. Having the tannin 
measure available would enable the ability to create a full model of the 
balance traits. The price data would be an interesting data point to 
investigate, is the pricing vs quality normally distributed, or does it shift 
towards higher? It would also be interesting to see a dataset with more 
varieties of wine, to get a more broad understanding of how the features affect 
quality, and how the features differ between wine varieties.


# References
\#many R related inqueries:  
http://www.cookbook-r.com/

\#read up on factor variables:  
https://www.stat.berkeley.edu/classes/s133/factors.html

\#read up on ggplot geom_density plots:  
https://www.r-bloggers.com/density-plot-with-ggplot/
http://ggplot2.tidyverse.org/reference/geom_density.html

\#read up on boxplot techniques:  
http://ggplot2.tidyverse.org/reference/geom_boxplot.html

\#correlation plot techniques:  
http://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram

\#plotting lines and shapes help:  
http://www.cookbook-r.com/Graphs/Shapes_and_line_types/

\#how to discern wine quality:  
http://www.dummies.com/food-drink/drinks/wine/how-to-discern-wine-quality/

\#how to measure tannins in wine:  
https://www.awri.com.au/wp-content/uploads/2014/01/measuring-wine-tannins.pdf

\#looked into citric acid in wine:  
http://www.livestrong.com/article/189520-what-drinks-do-not-contain-citric-acid/
