# Red Wine Quality Exploration  
## Bart Leatham  
### 04/18/2017

Wine experts judge wines on a variety of categories in an attempt to quantify a
wines quality.  One of the categories used is the 'balance' of a wine.  The 
balance consists of 4 measureable traits, sweetness, acidity, tannin and 
alcohol content. While personal taste will always prevail over a wine 'experts' 
opinion, this investigation into a [dataset](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1492550811255000&usg=AFQjCNGNw9JZHE_6c9hEJDIGpoLyC5MSkA) 
containing the judgments of at least three wine experts will attempt to 
quantify the relationships between the traits of a balanced wine and the 
perceived quality rating by the judges. This dataset consists of data related 
to red Portuguese "Vinho Verde" wine.  

#### Source of the dataset  
Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, 
Telmo Matos and Jose Reis (CVRVV) @ 2009  

#### Attribute Information  
   1 - fixed acidity (tartaric acid - g / dm^3)  
   2 - volatile acidity (acetic acid - g / dm^3)  
   3 - citric acid (g / dm^3)  
   4 - residual sugar (g / dm^3)  
   5 - chlorides (sodium chloride - g / dm^3  
   6 - free sulfur dioxide (mg / dm^3)  
   7 - total sulfur dioxide (mg / dm^3)  
   8 - density (g / cm^3)  
   9 - pH  
   10 - sulphates (potassium sulphate - g / dm3)  
   11 - alcohol (% by volume)  
   Output variable (based on sensory data):   
   12 - quality (score between 0 and 10)  
   
#### Attribute Descriptions:  
   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
   (do not evaporate readily)  

   2 - volatile acidity: the amount of acetic acid in wine, which at too high 
   of levels can lead to an unpleasant, vinegar taste  

   3 - citric acid: found in small quantities, citric acid can add 'freshness' 
   and flavor to wines  

   4 - residual sugar: the amount of sugar remaining after fermentation stops, 
   it's rare to find wines with less than 1 gram/liter and wines with greater 
   than 45 grams/liter are considered sweet  

   5 - chlorides: the amount of salt in the wine  

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
   growth and the oxidation of wine  

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
   concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine  

   8 - density: the density of water is close to that of water depending on the 
   percent alcohol and sugar content  

   9 - pH: describes how acidic or basic a wine is on a scale from 
   0 (very acidic) to 14 (very basic); most wines are between 3-4 on the 
   pH scale  

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
   (S02) levels, wich acts as an antimicrobial and antioxidant  

   11 - alcohol: the percent alcohol content of the wine  

   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)  

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

#install.packages("ggplot2", dependencies = T) 
#install.packages("knitr", dependencies = T)
#install.packages("dplyr", dependencies = T)
#install.packages("Hmisc", dependencies = T)
#install.packages("GGally", dependencies = T)
#install.packages("memisc", dependencies = T)
#install.packages("corrplot")


library(ggplot2)
library(dplyr)
library(knitr)
library(GGally)
library(memisc)
library(corrplot)
library(gridExtra)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_data}
# Load the Data
wine_df <- read.csv('wineQualityReds.csv')

#remove the 'X' column
wine_df$X <- NULL

#create correlation matrix
wine_corr <- cor(wine_df)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Define_Functions}

#This is a sandbox for functions to be called throughout EDA

#function to create histogram of feature count
feature_hist <- function(feature, data, binwidth = diff(range(data[column]))/30) {
    return(print(ggplot(aes_string(x = feature), data = data) + 
             geom_bar()))
  }

#function to create density plot of feature vs. quality_ranking
density_plot <- function(feature, data) {
  return(print(
    ggplot(aes_string(x = feature), data = data) +
      geom_density(aes(fill = Quality_Ranking), alpha = .3)
  ))
}

#function to create scatter plot of feature vs. quality_ranking
scatter_plot <- function(feature, data) {
  return(print(
    ggplot(aes_string(x = feature, y = "quality.bucket"), data = data) +
      geom_point(aes(fill = Quality_Ranking, color = Quality_Ranking ))
  ))
}

#function to create bar chart of feature vs quality
quality_bar <- function(feature, data) {
    return(print(ggplot(aes_string(x = "quality.bucket", y = feature), 
                        data = data) +
  geom_bar(stat = 'summary', fun.y = mean)
    ))
}

#function to create box plot of feature vs quality
quality_box <- function(feature, data) {
  return(print(ggplot(aes_string(x = "Quality_Ranking", y = feature), 
                        data = data) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1)))
}
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Cleaning}
#create new quality.bucket with discrete "low", "mid", "high" buckets
wine_df$quality.bucket <- cut(wine_df$quality, c(3, 4, 6, 8), 
                              labels = c("low", "mid", "high"), 
                              include.lowest = TRUE)

#convert quality to a factor type
wine_df$quality <- factor(wine_df$quality)

#assign above to a variable for cleaner plotting
Quality_Ranking <- wine_df$quality.bucket

```
# Overview of the data
```{r echo=FALSE, message=FALSE, warning=FALSE, Structure}
str(wine_df)
summary(wine_df)
```
I converted the 'quality' feature to a factor as it is discrete values, and 
created a 'quality.bucket' variable with low, mid and high quality buckets for 
more meaningful analysis.  
It should be noted, as shown in the summary above, that there are many more 
datapoints falling in the 'mid'(1319) quality bucket than 'low'(63) or 
'high'(217).  This should be considered when making any assumptions about the 
analysis.

# Univariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE}
for (name in names(wine_df)) {
      feature_hist(name, wine_df)
  }
```


# Univariate Analysis

### What is the structure of your dataset?
1599 obs. of  13 variables.  11 input variables, 2 output variables.

### What is/are the main feature(s) of interest in your dataset?
This analysis is focused on the balance of the wines, of which the traits are 
sweetness, acidity, tannin and alcohol. The dataset does not contain a measure
for tannins, so the investigation will focus on sweetness 
(measured by residual.sugar), acidity (measured by pH), the alcohol content
and quality rating.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I am curious how the density of the wine will relate to the quality rating.

### Did you create any new variables from existing variables in the dataset?
I bucketed the quality ratings into a quality.bucket variable, with low, med, 
high being the buckets.  This allows for a more meaningful analysis, as the 
counts for some quality rankings were low compared to others.  

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?  
This dataset met tidy data requirements as delivered, so the only cleaning I 
performed was to remove the X column as it was just a line count index.  
Most of the features have an overall normal distribution, some skewed to the 
right, with a long tail of outliers.

# Bivariate Plots Section

### Correlation data
Below we see the correlation values between all input and output variables, as 
well as a graphical representation.
```{r echo=FALSE, message=FALSE, warning=FALSE, Correlation}
#summary(wine_df)
head(round(wine_corr, 2), 13)
corrplot(wine_corr, method="circle", type="lower")
```

A quick look at the correlation between the variables shows that there are no 
strong correlations between quality and the various input variables, the 
strongest postive correlation being the alcohol content at r= 0.48, and the
strongest negative correlation being the volatile acidity at r= -0.39.


### Bivariate Plots
After perfoming multiple plotting techniques (hist, bar, density, scatter, box) 
for introductory analysis, I found that using a box plot of each variable
vs. quality.bucket yielded the most useful visual data for Bivariate analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
#iterate over features in wine_df and create initial exploratory plots 
for (name in names(wine_df)) {
    if (name != 'quality' & name != 'quality.bucket' ) {
      quality_box(name, wine_df)
    }
  }
```

## Bivariate Analysis
Upon analyzing the above plots, we can see that the higher the chloride and 
volatile.acidity, the lower the quality.  
The higher the fixed.acidity, sulfates, alcohol and citric.acid, the higher 
the quality.  
It is interesting to note, that if relying on the box plots alone, 
one could infer that the variables mentioned could have a large correlation to 
the quality of the wine. Since we have looked at the correlation values 
explicity, we know that this is not case, there is a correlation, though not a
strong one.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
Of the 4 features of interest, density and alcohol content have the strongest
correlations to the quality of the wine. Density has a r = -0.17 while alcohol
has r = 0.48 correlation to the quality of the wine. The pH and residual sugar
have low correlations (r = -0.06 and r = 0.01 respectively).  I expected
that the residual sugar would have more of an impact, as I often hear wines 
described as being too sweet.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
It was interesting to see that the correlation data reinforced some 
relationships that would be expected.  For instance, the various *.acid 
features have relatively strong correlations to the pH.

### What was the strongest relationship you found?
The strongest relationship that I found was between free.sulfur.dioxide and
total.sulfur.dioxide.  This makes sense, as the free.sulfur.dioxide is a subset
of the total.sulfur.dioxide, so it is not particularly interesting.

# Multivariate Plots Section
I was interested in looking deeper into the relationships between the residual 
sugar, density, pH, alcohol content and quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, GGpairs}
ggpairs(wine_df, columns = c(4,8,9,11,13), 
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'))), width =6,
        height = 6)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(aes_string(x = 'residual.sugar', y = 'alcohol'), data = wine_df, width =6,
        height = 16) +
  geom_smooth(aes(fill = Quality_Ranking, color = Quality_Ranking))

p2<- ggplot(aes_string(x = 'pH', y = 'alcohol'), data = wine_df, width =6,
        height = 16) +
  geom_smooth(aes(fill = Quality_Ranking, color = Quality_Ranking))

p3<- ggplot(aes_string(x = 'density', y = 'alcohol'), data = wine_df, width =6,
        height = 16) +
  geom_smooth(aes(fill = Quality_Ranking, color = Quality_Ranking) )

grid.arrange(p1, p2, p3, ncol = 1)

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Above we can see that higher alcohol levels consistently yield higher quality 
ranking results.  
For residual sugar, we can see that there are no high quality ranked wines with
a residual sugar level over ~9g/dm^3, giving credence to some wines being too
sweet.  
For the pH, we can see that above ~3.75 and below ~2.5 there are no high 
quality ranked wines, leading me to believe that a pH of 2.5-3.75 is the
range for a well balanced wine, not too acidic or basic.  It can be noted
though, that on the pH scale, that range is quite acidic.  
For the density of the wine, we cannot infer much as to quality, we do see
correlation to the fact that lower density results in higher alcohol levels, 
which we have seen have a moderate correlation to quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
p1<- ggplot(aes(x = alcohol, fill = Quality_Ranking, color = Quality_Ranking), 
            data = wine_df) +
        ggtitle("Wine Quality as a function of Alcohol Content") +
        geom_density(alpha = 0.4) +
        labs(x = "Alcohol (% by volume)", y = "probablity density")

p2<- ggplot(aes(x = pH, fill = Quality_Ranking, color = Quality_Ranking), data = wine_df) +
        ggtitle("Wine Quality as a function of pH measure") +
        geom_density(alpha = 0.4) +
        labs(x = "pH (measure of acidity)", y = "probablity density")

p3<- ggplot(aes(x = residual.sugar, fill = Quality_Ranking, color = Quality_Ranking), 
            data = wine_df) +
        ggtitle("Wine Quality as a function of Residual Sugar Content") +
        geom_density(alpha = 0.4) +
        labs(x = "Residual Sugar (g/dm^3)", y = "probablity density") +
  scale_x_continuous(limits = c(0,4))

grid.arrange(p1, p2, p3, ncol = 1)
```

### Description One
The above plots shows the probability density of each of our traits of interest
in investigating the 'balance' of our wine observations.  
We can see that in isolation, higher alcohol content, lower pH and slightly 
higher residual sugar lead to higher quality wines.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = alcohol/pH, fill = Quality_Ranking, color = Quality_Ranking),
        data = wine_df) +
        ggtitle("Wine Quality as a function of Alcohol/pH ratio") +
        geom_density(alpha = 0.4) +
        labs(x = "Alcohol/pH ratio", y = "probablity density")

```

### Description Two
This plot shows the relationship between the quality of a wine and the ratio of
alcohol/pH.  It shows the optimum ratio of alcohol to pH as being 3.5, with a 
clear differentiation from the optimum ratios for low or mid ranked wines.  
This reinforces the fact that wines with a lower pH, and a higher alcohol 
content are ranked higher in quality in the given dataset.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = density, fill = Quality_Ranking, color = Quality_Ranking),
        data = wine_df) +
        ggtitle("Wine Quality as a function of Density") +
        geom_density(alpha = 0.4) +
        labs(x = "Density (g/cm^3)", y = "probablity density")
```

### Description Three
This last plot adresses my additional feature of interest, does the density of
the wine effect the quality of the wine? From the plot above we can see that
lower density does correlate with a higher quality rating, confirmed by the 
r = -0.17 found in the correlation section of this investigation.  With density
having an r = 0.5 correlation to alcohol content, common sense is validated
knowing that alcohol has a lower density than water (1g/cm^3), so higher 
alcohol content lowers the density of the wine.

# Reflection
This red wine dataset provided 1599 observations of 11 input and 1 output 
variable.  I first looked into the distribution of the variable values, then 
moved on to their correlations.  Once I saw that the correlations were 
generally weak, especially with regard to the reported quality rating of the 
wine, I realized that I needed to come up with a direction to take the 
investigation.  At that point I researched how wine quality is generally 
defined by experts, and decided to pursue the measure of 'Balance' for a wine
and investigated its traits as much as possible with the data provided.  
The main struggle I faced in this investigation was the fact that none of the
independant variables were statistically strongly correlated to the quality
ranking of the wines.  
I was surprised to see that the features were not more strongly correlated to
the quality rating of the wines.  Before this investigation I would have 
guessed that alcohol, pH and sweetness all would have been more highly 
correlated with the quality rating. To me this proves that wine quality ratings
are highly subjective, and a wine consumer would do best to trust their
tastebuds (and wallets) over being overly concerned with how a wine expert
views their choice.
Future work with this dataset I would be interested in having both a measure
of tannins, and the price of the wine included. Having the tannin measure
available would enable the ability to create a full model of the 'Balance' 
traits.  The price data would be an interesting data point to investigate, is
the pricing vs quality normally distributed, or does it shift towards higher? 
It would also be interesting to see a dataset with more varieties of wine, to 
get a more broad understanding of how the features effect quality, and how the
features differ between wine varieties.


# References
\#many R related inqueries:  
http://www.cookbook-r.com/

\#read up on factor variables:  
https://www.stat.berkeley.edu/classes/s133/factors.html

\#read up on ggplot geom_density plots:  
https://www.r-bloggers.com/density-plot-with-ggplot/
http://ggplot2.tidyverse.org/reference/geom_density.html

\#read up on boxplot techniques:  
http://ggplot2.tidyverse.org/reference/geom_boxplot.html

\#correlation plot techniques:  
http://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram

\#plotting lines and shapes help:  
http://www.cookbook-r.com/Graphs/Shapes_and_line_types/

\#how to discern wine quality:  
http://www.dummies.com/food-drink/drinks/wine/how-to-discern-wine-quality/

\#how to measure tannins in wine:  
https://www.awri.com.au/wp-content/uploads/2014/01/measuring-wine-tannins.pdf
